A Core Calculus of Metaclasses

### 2004-12-12 metaclasses/paper.tex
lots of paper updates

\section{Introduction}
One of the stated benefits of object-oriented languages is their
ability to model aspects of the world in the object hierarchy.
However, most such languages are unable to model many simple and
common relationships. For example, consider the relationship between
physical quantities, units of measurement, and physical dimensions. It
is obvious that ``length'' is a dimension, and that ``3 feet'' is a
length.  However, modeling this relationship in a conventional
language such as the Java Programming Language is difficult without
attributing to some of these concepts properties which do not properly
belong to them.  It is natural to model {\txt 3 feet} as an instance
of a class {\txt Length}. But if we were to define a class {\txt
Dimension} and define {\txt Length} to be an instance of class {\txt
Dimension}, then {\txt Length} could not be a class and so {\txt 3
feet} could not be an instance of {\txt Length}. Alternatively, we
might define {\txt Length} to be a subclass of {\txt Dimension}. But
then we still cannot define {\txt 3 feet} to be an instance of {\txt
Length} because anything that is an instance of {\txt Length} would
also be an instance of class {\txt Dimension} and {\txt 3 feet} is
obviously not a dimension. The key limitation preventing us from
modeling this domain properly is that we cannot define a single
concept to serve both as a class and as an instance of a class.

Another example is provided by Welty and Ferrucci in \cite{FW96}, where
they show the difficulties in modeling a simple
ontology including the notions {\txt Species} and {\txt Eagle}, while
also capturing the relationship between {\txt Eagle} and {\txt Harry},
a particular eagle.  One solution they consider and reject is to
separate the notion of ``eagle'' into two concepts: {\txt the Eagle}
and {\txt Eagle}, with the former an instance of {\txt Species} and
the latter a class whose instances are specific birds.  This approach
is inadequate because it fails to express a relationship between the
notions {\txt the Eagle} and {\txt Eagle}.
Again, the limitation is that we cannot define a class as an instance of another class.
%Welty and Ferrucci proceed to explain how this limitation prevents conventional modeling
%languages from capturing many important aspects of the world.  
%The languages Ferrucci and Welty discuss cannot accommodate multiple
%universes of discourse, in which a single thing may have different meanings in different universes.

@@

A more expressive metaclass system in ObjVlisp, described informally
by Conte in \cite{objvlisp}.  In ObjVlisp, there is no restriction on
the number of instances of a metaclass, and metaclasses are not
required to share a common superclass. The instance methods of a
metaclass are inherited as the static members of its instances.  Conte
makes a number of arguments for the benefits of generalized
metaclasses.  In response to complaints that the Smalltalk80 metaclass
system is too complex, he suggests that a more powerful and flexible
system has the potential to be much simpler.
%, arguing that ``The ObjVlisp model's primary advantage is uniformity.''
However, ObjVlisp does not include a static type system. 
%Cointe quotes from \cite{cardelli:multiple}: ``With respect to
%Simula, Smalltalk also abandons ... strong typing, allowing it to
%... introduce the notion of meta-classes''.   
In this paper, we show that this limitation is not necessary. We
present a semantics for a statically typed calculus of metaclasses
where, as in ObjVlisp, every class can serve as both an instance of a
class and as a class of instances, and there is no limitation on the
levels of nesting of metaclasses. In the process, we also reveal
several interesting properties about the structure that a nominal type
system for metaclasses must take. We then present a formal operational
semantics for this calculus, as well as a proof of type soundness.

@@

MCJ is a generically-typed object-oriented calculus based on
Featherweight GJ that includes metaclasses. An MCJ program consists of
a sequence of class definitions followed by a trailing
expression. This trailing expression is evaluated in the context of
the class table induced by the class definitions.  Before discussing
the formal specification of MCJ, we first give a high-level overview
of the nature of MCJ class definitions.

Each class is an instance of a \emph{metaclass}.  A metaclass is
either a user-defined class or class {\txt Object}.  We say that if a
class {\txt C} is an instance of a class {\txt D}, then {\txt C} is an
\emph{instance class} of {\txt D}, and that {\txt D} is the
\emph{immediate containing class}, or \emph{kind}, of {\txt C}.  If
class {\txt E} is a superclass of {\txt D} then we also say that {\txt
C} is an instance class of {\txt E} and that {\txt E} is a containing
class of {\txt C}.

@@

A class definition consists of a header plus a collection of fields
and methods associated with the class.  The header names the class and
specifies the type parameters, as well as the superclass and the kind.
For example, the following header declares a class {\txt C} with
superclass {\txt Object} and kind {\txt D} and no type parameters:

\begin{verbatim}
class C kind D extends Object {...}
\end{verbatim}

Each class may have both class members and instance members. Class
members define the behavior of the class when used as an
instance. Instance members define the behavior of instances of a
class.  In the abstract syntax presented in this paper, class members
are distinguished from instance members by order, and with the keyword
{\txt class} \footnote{This syntax is not ambiguous because MCJ has no
inner classes.}.  Members of a class definition are laid out as
follows: first, class fields are defined, followed by class methods,
then instance fields and finally instance methods.  For example, the
following class definition includes a class method {\txt m} and an
instance method {\txt n}:

\begin{verbatim}
class C kind Object extends Object {
   class Object m() {...}
   Object n() {...}
}
\end{verbatim}

Class fields and methods bear a resemblance to static fields and
methods, but there are important differences. For a given class {\txt
C}, instance members of {\txt C}'s kind are inherited as class
members of {\txt C}. For example, a class method {\txt m} in {\txt C}
with the same name as an instance method of {\txt C}'s kind
overrides the definition of {\txt m} in the kind (and it must have the
same signature as the overridden method). Note that if an instance
class {\txt C} of type {\txt T} is assigned to a variable {\txt v} of
type {\txt T}, references to the instance members of {\txt v} refer to
the class members of {\txt C}.  As in \cite{DimUnits}, a class is
allowed to define both a static method and an instance method of the
same name. We prevent ambiguity in member references by requiring that
all references to class members in MCJ must explicitly denote the
receiver.

@@

\begin{verbatim}
class A kind Object extends Object {...}
class B kind Object extends A {...}
class C kind B extends Object {...}
\end{verbatim}

@@

The generic type system of MCJ is similar to the generic type system
in \FGJ. However, as in \cite{NextGen, DimUnits}, we allow for type
variables to occur in type-dependent contexts such as casts,
preventing the use of type erasure as an implementation technique. In
addition, the receiver of a reference to a class member may be a type
variable. Unlike the system presented in \cite{MixGen}, MCJ does not
support first-class genericity because a naked type variable must not
appear in the {\txt extends} clause or {\txt kind} clause of a class
definition and {\txt new} expressions on naked type variables are not
supported. Also, polymorphic methods (which are orthogonal to the
features we explore) are not supported.

@@

\begin{verbatim}
class C<T extends (D,E)>
   kind Object
   extends Object {...}
\end{verbatim}

@@

One benefit of a metaclass system is that constructors no longer need
to play such a central role in the language. Class instances exist at
the beginning of a program execution; they need not be
constructed. Non-class instances are naturally constructed with class
(factory) methods \cite{GOF}; in this way a class instance can be
passed as a type parameter and factory methods can be called on the
class instance to construct new instances without stipulating the
class of the constructed instance. This flexibility is similar to that
provided with first-class genericity \cite{NextGen, MixGen}, but
obviates the need for {\txt with} clauses because the bound on the
kind of a type variable stipulates what factory methods can be called
on it.  A class can define class methods with arbitrary signatures
that return new instances of the class, but each such method must
ultimately include a {\txt new} expression, which can occur only
within the scope of the class whose instance it returns. A {\txt new}
expression looks like a call to method named {\txt new}, that takes a
single argument for each type parameter and each instance field of the
syntactically enclosing class. The result of a {\txt new} expression
is a new instance of the enclosing class whose type parameters are
instantiated with, and whose fields are initialized with, the given
arguments. In our formal semantics, {\txt new} expressions are
annotated with the name of the enclosing class. These annotations need
not be added by a programmer; they could be added easily by a
straightforward syntactic preprocessing over the program, since they
can be determined merely by the lexical scopes of the {\txt new}
expressions.

@@

% In a {\txt new} expression, all instance fields must be initialized at
% creation time. This requirement eliminates the problem of
% uninitialized instance fields. However, when a class is the kind of
% another class, the instance fields of the kind become class fields of
% the instance class.  Since the instance class can then be used as a
% value with no further initialization, these variables must be
% initialized somewhere. Our solution is to require that every class
% field inherited from a container class be included in a new class
% field definition and that every class field in a class definition
% include an initializing expression.

@@

Given that class references can be used as expressions, it is natural
to ask: what is the type of a class reference?  
%In the world of the
%$\lambda$-calculus, such types are known as kinds.  
In MCJ, however, the kind of a class does not capture all of the
properties of the class as a value.  For example, a class may add new
class fields or class methods which are not present in the kind.
Therefore, each class freely generates a new type, which is the type
of the class considered as a value.  We represent this type with the
{\txt typeOf} operator, which takes a class instantiation and produces
a type that is not also a class.  Because this type is not a class, it
cannot be the superclass or kind of another class, and it cannot serve
as the instantiation of a generic type parameter.

@@

\begin{verbatim}
class A() kind Object extends Object {...}
class B() kind A extends Object {...}
\end{verbatim}

induce the following type relationships:
\begin{displaymath}
\begin{array}{c}
\typ \B {\typeof \B} \\
\subtype {\typeof \B} \A \\
\B~ {\txt instanceof}\ \A
\end{array}
\end{displaymath}

\hspace{-.43cm}where we use the convention that {\txt A:B} denotes
that {\txt A} has type {\txt B}, and {\txt A<:B} denotes that {\txt A}
is a subtype of {\txt B}.

The name {\txt typeOf} has been used in other contexts to refer to a
runtime operation that determines the dynamic type of a object.
However, despite the name similarity, this function bears no
relationship to the operator described here.  In MCJ, an application
of {\txt typeOf} is a static type reference.

@@

The standard subclassing relationship, like subtyping, is transitive.
That is, if $\B$ is a subclass of $\C$ and $\C$ is a subclass of $\D$,
then $\B$ is a subclass of $\D$.  However, this relationship does not
hold for {\txt instanceof} relationships.  In general, if $\B$ is an
instance of $\C$ and $\C$ is an instance of $\D$, no judgment
about the relationship of $\B$ to $\D$ can be inferred.

Another difference between subclassing and {\txt instanceof}
relationships is that cycles can occur among instance class
relationships. The simplest such cycle is:

\begin{displaymath}
\typeFront {\typeArg \C{}} {\typeArg \C {}} \Object \braces{}
\end{displaymath}

Although this pathology appears to be dangerously close to Russell's
paradox, it does not lead to inconsistencies.  Class \C\ is an
instance of itself. It can be used in any context where one of its
instances can be used.  \C\ can be thought of as a self-replicating
value. This class hierarchy causes no problems for method lookup
because method resolution never proceeds through more than a single
containing class to an instance class. Therefore, we see no reason to
disallow it.

@@

\begin{verbatim}
class Dimension kind Object extends Object { ... }
class Length kind Dimension extends Object { ... }
class Meter kind Length extends Object { ... }
\end{verbatim}

Here {\txt Meter} is a singleton class (there is only one Meter, in
the Platonic sense).  We have easily expressed the desired type
relationships, and can statically check program invariants that rely
on dimensional relationships.

@@

\begin{verbatim}
class Species kind Object extends Object { ... }
class Eagle kind Species extends Object {
      class Eagle make(String name) { ... }
}
Eagle harry = Eagle.make("Harry")
\end{verbatim}

@@

\begin{verbatim}
class Species extends Object { ... }
class TheEagle extends Species { ... }
class Eagle extends Object { ... }
Eagle harry = new Eagle("Harry")
\end{verbatim}

There are at least two substantial problems with this code.  First,
{\txt TheEagle} and {\txt Eagle} have no relationship to each other in
the type system.  One is a singleton class, representing a particular
species, and one is a name for a set of objects, being the members of
that species.  The fundamental nature of species has a two-level
containment relationship, which the type system fails to express.
This means that our type checker cannot determine what we want and
cannot help us avoid mistakes.

Another problem, which is in some ways just a symptom of the first
problem but which bears special attention, is the use of generic
types.  Imagine the following method, where T is a type parameter in
scope:\footnote{Generic methods would be straightforward to add to
MCJ, but we exclude them to simplify the presentation of the type
system and to focus on the important interactions of metaclasses with
other language features.}

\begin{verbatim}
foo(T x) { ... }
\end{verbatim}

If we want to perform the call {\txt foo(harry)}, then the bound on
{\txt T} must be either {\txt Object}, in which case {\txt foo} must
have no knowledge of the object, or Eagle, in which case the function
cannot handle multiple {\txt Species}.  The ``solution'' in
conventional languages is to use a bound of {\txt Object} and insert a
downcast, which fails at runtime if the wrong argument is passed.  In
MCJ {\txt T} can be bound by kind {\txt Species}, and instantiated
with {\txt Eagle}, allowing the original call to be type-checked.
This gives the programmer both safety and expressiveness.

@@

One of the advantages of our metaclass framework is that it allows
clean expression of several common design patterns in the language,
rather than requiring that they be expressed with abstraction
techniques on top of the language.  For example, a number of the
classic design patterns deal with object creation, as discussed above
in section \ref{privconst}.  It is clear that the concept of a Factory
method is easily (and necessarily) expressed in MCJ.  However, a
number of other patterns also find easy expression.  The Abstract
Factory pattern is a simple application of inheritance in the
metaclass hierarchy.  The Prototype pattern is, as mentioned in the
Design Patterns book, supported in languages with metaclasses.
Finally, there is no need for the Singleton pattern; a class itself
can serve as a singleton.  While classes can be used as limited
singletons in the Java Programming Language and in \Cpp, doing so is
undesirable because the static behavior of a class cannot inherit from
another class.  In fact, the inflexibility of class or static
operations is one of the rationales cited for the Singleton pattern.

@@

A number of symbols are used to abbreviate keywords: $\superSym$
stands for {\txt extends} and $\kindSym$ stands for {\txt kind}. Also
$@$ represents concatenation of sequences of syntactic constructs.
Finally, $CT(\T)$ is a lookup in the class table for the definition of
$\T$.

A number of restrictions on MCJ programs are implicit in the formal
rules.  First, we assume that all sequences of methods and fields are
free of duplicates.  Second, there is an implicit well-formedness
constraint on programs that no class be a superclass of itself, either
directly or indirectly (however, as discussed below, cycles in the
kind hierarchy are allowed).  Third, we assume that {\txt this} is
never used as the name of a variable, method or field.  We also take
\Object\ to be a distinguished member of the hierarchy, with no
specific definition, which does not have a superclass or a kind.

@@

This is a class named $\C$, with no type arguments, with kind
$\Object$ and superclass $\Object$, which has no fields or methods
whatsoever.  In examples that follow, we omit empty {\txt <>}.  Note
that both the kind and the superclass are $\Object$.  MCJ does not
have a distinguished class $\Class$.  Interestingly, we determined
that a class $\Class$ would need no consideration from the type
system.  Because both $\Class$ and $\Object$ would sit atop the
hierarchy with no contents, there is no need to include
both of them.  In a more substantial language, where $\Object$ might
have methods or fields, there might be a use for a distinguished
$\Class$ class, but it would not need any special treatment by
the type system.

A more complicated MCJ class is:\footnote{Here we use {\txt extends}
for $\superSym$ and {\txt kind} for $\kindSym$.}

\begin{verbatim}
class Pair<A extends (Object, Object),
           B extends (Object, Object)>
   kind Object extends Object
{
 class Pair<A,B> make(A a, B b){ new(A,B) (a,b) } ;
 A fst;
 B snd;
 Pair<B,B> setfst(B b){
     return new <B,B> (b, this.snd);
 };
}
\end{verbatim}

This class specifies a polymorphic pair data structure.  It also
demonstrates a number of important MCJ features.  First, we see two
different kinds of methods.  The {\txt make} method creates a new
instance of {\txt Pair}.  This method contains a {\txt new}
expression, which initializes the instance fields of the class
positionally, so that the first argument to {\txt make} becomes {\txt
 fst}, and the second becomes {\txt snd}.  Also, note that we must
mention the type parameters of the newly created instance.

Second, we have an instance method, {\txt setfst}, which creates a new
pair with the old second element, and a new first element which has
the same type as the second element.

Finally, adding the following to the above definition of {\txt Pair}
gives a full MCJ program:

\begin{verbatim}
class O kind Object extends Object
{
 class O make(){return new () ()};
}

@@

This program, which creates two instances of {\txt O}, then creates
a {\txt Pair} to hold them both, and finally selects one of the two to
become the value of the program, demonstrates three of the expression
forms in MCJ.  {\txt Pair<O,O>} is a type, which, by itself, can
be a value.  {\txt O.make()} is a method invocation (here with a
class, not an instance, as the receiver).  The entire expression is a
field access, of the {\txt fst} instance field.

>>>>

\section{Introduction}
One of the stated benefits of object-oriented languages is their
ability to model aspects of the world in the object hierarchy.
However, most such languages are unable to model many simple and
common relationships.

An excellent example is provided by Welty and Ferrucci in \cite{FW96}, where
they show the difficulties in modeling a simple
ontology including the notions {\txt{Species}} and {\txt{Eagle}}, while
also capturing the relationship between {\txt{Eagle}} and {\txt{Harry}},
a particular eagle.

To see the problem, consider a system with a class \txt{Eagle} and an
instance \txt{Harry}.  The static type system can ensure that any
messages that \txt{Harry} responds to are also understood by any other
instance of \txt{Eagle}.

However, even in a language where classes themselves can have methods,
such relationships cannot be expressed.  For example, in a language
with static methods, we can write the expression
\txt{Eagle.isEndangered()}

However, if \txt{Salmon} is a species, our type system provides us no
way of requiring that it also have an \txt{isEndangered} method.  The
solution that Ferruci and Welty suggest, and that we develop in this
paper, is to allow a consistent treatment of both instances such as
\txt{Harry} and classes such as \txt{Eagle}.

One solution they consider and reject is to separate the notion of
``eagle'' into two concepts: {\txt{the Eagle}} and {\txt{Eagle}}, with
the former an instance of {\txt{Species}} and the latter a class whose
instances are specific birds.  This approach is inadequate because it
fails to express a relationship between the notions {\txt{the Eagle}}
and {\txt{Eagle}}.  Therefore, if we create another species such as
\txt{Salmon} in our universe, we must define one new class and one new
instance.  This adds new invariants which the programmer must
maintain, with no help from the type system.  Instead, we propose that
the type system should allow the programmer to express the desired
relationship directly, and enforce the invariants naturally.

For example, consider the relationship between
physical quantities, units of measurement, and physical dimensions. It
is obvious that ``length'' is a dimension, and that ``3 feet'' is a
length.  However, modeling this relationship in a conventional
language such as the Java Programming Language is difficult without
attributing to some of these concepts properties which do not properly
belong to them.  It is natural to model {\txt{3 feet}} as an instance
of a class {\txt{Length}}. But if we were to define a class {\txt
Dimension} and define {\txt{Length}} to be an instance of class {\txt
Dimension}, then {\txt{Length}} could not be a class and so {\txt 3
feet} could not be an instance of {\txt{Length}}. Alternatively, we
might define {\txt{Length}} to be a subclass of {\txt{Dimension}}. But
then we still cannot define {\txt 3 feet} to be an instance of {\txt
Length} because anything that is an instance of {\txt{Length}} would
also be an instance of class {\txt{Dimension}} and {\txt 3 feet} is
obviously not a dimension. The key limitation preventing us from
modeling this domain properly is that we cannot define a single
concept to serve both as a class and as an instance of a class.

Again, the limitation is that we cannot define a class as an instance of another class.
%Welty and Ferrucci proceed to explain how this limitation prevents conventional modeling
%languages from capturing many important aspects of the world.  
%The languages Ferrucci and Welty discuss cannot accommodate multiple
%universes of discourse, in which a single thing may have different meanings in different universes.

@@

A more expressive metaclass system in ObjVlisp, described informally
by Cointe in \cite{objvlisp}.  In ObjVlisp, there is no restriction on
the number of instances of a metaclass, and metaclasses are not
required to share a common superclass. The instance methods of a
metaclass are inherited as the static members of its instances.  Conte
makes a number of arguments for the benefits of generalized
metaclasses.  In response to complaints that the Smalltalk80 metaclass
system is too complex, he suggests that a more powerful and flexible
system has the potential to be much simpler.
%, arguing that ``The ObjVlisp model's primary advantage is uniformity.''
However, ObjVlisp does not include a static type system. 
%Cointe quotes from \cite{cardelli:multiple}: ``With respect to
%Simula, Smalltalk also abandons ... strong typing, allowing it to
%... introduce the notion of meta-classes''.   
In this paper, we show that this limitation is not necessary. We
present a semantics for a statically typed calculus of metaclasses
where, as in ObjVlisp, every class can serve as both an instance of a
class and as a class of instances, and there is no limitation on the
levels of nesting of metaclasses. In the process, we also reveal
several interesting properties about the structure that a nominal type
system for metaclasses must take. We then present a formal operational
semantics for this calculus, as well as a proof of type soundness.

@@

MCJ is a object-oriented calculus with generic types based on
Featherweight GJ which includes metaclasses. An MCJ program consists of
a sequence of class definitions followed by a trailing
expression. This trailing expression is evaluated in the context of
the class table induced by the class definitions.  Before discussing
the formal specification of MCJ, we first give a high-level overview
of the nature of MCJ class definitions.

Each class is an instance of a \emph{metaclass}.  A metaclass is
either a user-defined class or class {\txt{Object}}.  We say that if a
class {\txt C} is an instance of a class {\txt D}, then {\txt C} is an
\emph{instance class} of {\txt D}, and that {\txt D} is the
\emph{immediate containing class}, or \emph{kind}, of {\txt C}.  If
class {\txt E} is a superclass of {\txt D} then we also say that {\txt
C} is an instance class of {\txt E} and that {\txt E} is a containing
class of {\txt C}.

@@

A class definition consists of a header plus a collection of fields
and methods associated with the class.  The header names the class and
specifies the type parameters, as well as the superclass and the kind.
For example, the following header declares a class {\txt C} with
superclass {\txt{Object}} and kind {\txt D} and no type parameters:

\begin{verbatim}
class C kind D extends Object {...}
\end{verbatim}
\footnote{In other examples, we will leave out the \kind\ and \extends\
declaration when they refer to \Object.}

Each class may have both class members and instance members. Class
members define the behavior of the class when used as an
instance. Instance members define the behavior of instances of a
class.  In the abstract syntax presented in this paper, class members
are distinguished from instance members by order, and with the keyword
\class \footnote{This syntax is not ambiguous because MCJ has no
inner classes.}.  Members of a class definition are laid out as
follows: first, class fields are defined, followed by class methods,
then instance fields and finally instance methods.  For example, the
following class definition includes a class method {\txt m} and an
instance method {\txt n}:

\begin{verbatim}
class C {
   class Object m() {...}
   Object n() {...}
}
\end{verbatim}

Class fields and methods bear a resemblance to static fields and
methods, but there are important differences. For a given class {\txt
C}, instance members of {\txt C}'s kind are inherited as class
members of {\txt C}. For example, a class method {\txt m} in {\txt C}
with the same name as an instance method of {\txt C}'s kind
overrides the definition of {\txt m} in the kind (and it must have the
same signature as the overridden method). Note that if an instance
class {\txt C} of type {\txt T} is assigned to a variable {\txt v} of
type {\txt T}, references to the instance members of {\txt v} refer to
the class members of {\txt C}.  As in \cite{DimUnits}, a class is
allowed to define both a class method and an instance method of the
same name. We prevent ambiguity in member references by requiring that
all references to class members in MCJ must explicitly denote the
receiver.

@@

\begin{verbatim}
class A {...}
class B extends A {...}
class C kind B {...}
\end{verbatim}

@@

The generic type system of MCJ is similar to the generic type system
in \FGJ. However, as in \cite{NextGen, DimUnits}, we allow for type
variables to occur in type-dependent contexts such as casts,
preventing the use of type erasure as an implementation technique. In
addition, the receiver of a reference to a class member may be a type
variable. Unlike the system presented in \cite{MixGen}, MCJ does not
support first-class genericity because a naked type variable must not
appear in the \extends\ clause or \kind\ clause of a class definition
and {\txt{new}} expressions on naked type variables are not
supported.\footnote{First-class genericity is difficult to integrate
with our presentation of constructors.} Also, polymorphic methods
(which are orthogonal to the features we explore) are not supported.

@@

\begin{verbatim}
class C<T extends D kind E> {...}
\end{verbatim}

@@

One benefit of a metaclass system is that constructors no longer need
to play such a central role in the language. Class instances exist at
the beginning of a program execution; they need not be
constructed. Non-class instances are naturally constructed with class
(factory) methods \cite{GOF}; in this way a class instance can be
passed as a type parameter and factory methods can be called on the
class instance to construct new instances without stipulating the
class of the constructed instance. This flexibility is similar to that
provided with first-class genericity \cite{NextGen, MixGen}, but
obviates the need for {\txt{with}} clauses because the bound on the
kind of a type variable stipulates what factory methods can be called
on it.

For example, in MixGen the following program is legal:

\begin{verbatim}
class C<T extends S with T()> {
  T foo() { return new T(); }
}
\end{verbatim}

However, to make this program compile, a \txt{with} clause is required
on the type parameter \txt{T}.  This is because the type associated
with a class does not describe its constructors.

In MCJ, this program can be naturally rewritten with a factory method.

\begin{verbatim}
class C<T extends S> {
  T foo() { return T.make(); }
}
\end{verbatim}

Here, no with clause is required.

A class can define class methods with arbitrary signatures that return
new instances of the class, but each such method must ultimately
include a {\txt{new}} expression, which can occur only within the
scope of the class whose instance it returns. A {\txt{new}} expression
looks like a call to method named {\txt{new}}, that takes a single
argument for each type parameter and each instance field of the
syntactically enclosing class. The result of a {\txt{new}} expression
is a new instance of the enclosing class whose type parameters are
instantiated with, and whose fields are initialized with, the given
arguments. In our formal semantics, {\txt{new}} expressions are
annotated with the name of the enclosing class. These annotations need
not be added by a programmer; they can be added easily by a
straightforward syntactic preprocessing over the program, since they
can be determined merely by the lexical scopes of the {\txt{new}}
expressions.

@@

% In a {\txt{new}} expression, all instance fields must be initialized at
% creation time. This requirement eliminates the problem of
% uninitialized instance fields. However, when a class is the kind of
% another class, the instance fields of the kind become class fields of
% the instance class.  Since the instance class can then be used as a
% value with no further initialization, these variables must be
% initialized somewhere. Our solution is to require that every class
% field inherited from a container class be included in a new class
% field definition and that every class field in a class definition
% include an initializing expression.

@@

Given that class references can be used as expressions, it is natural
to ask: what is the type of a class reference?  
%In the world of the
%$\lambda$-calculus, such types are known as kinds.  
In MCJ, however, the kind of a class does not capture all of the
properties of the class as a value.  For example, a class may add new
class fields or class methods which are not present in the kind.
Therefore, each class freely generates a new type, which is the type
of the class considered as a value.  We represent this type with the
{\txt{typeOf}} operator, which takes a class instantiation and produces
a type that is not also a class.  Because this type is not a class, it
cannot be the superclass or kind of another class, and it cannot serve
as the instantiation of a generic type parameter.  We refer to this as
{\txt{typeOf}} because it is the type of the expression that is its
argument.

@@

\begin{verbatim}
class A {...}
class B kind A {...}
\end{verbatim}

induce the following type relationships:
\begin{displaymath}
\begin{array}{c}
\typ \B {\typeof \B} \\
\subtype {\typeof \B} \A \\
\B~ {\mathtt {instanceof}}\ \A
\end{array}
\end{displaymath}

\hspace{-.43cm}where we use the convention that {\txt{A:B}} denotes
that {\txt A} has type {\txt B}, and {\txt{A<:B}} denotes that {\txt A}
is a subtype of {\txt B}.

The name {\txt{typeOf}} has been used in other contexts to refer to a
runtime operation that determines the dynamic type of a object.
However, despite the name similarity, this function bears no
relationship to the operator described here.  In MCJ, an application
of {\txt{typeOf}} is a static type reference.

@@

The standard subclassing relationship, like subtyping, is transitive.
That is, if $\B$ is a subclass of $\C$ and $\C$ is a subclass of $\D$,
then $\B$ is a subclass of $\D$.  However, this relationship does not
hold for {\txt{instanceof}} relationships.  In general, if $\B$ is an
instance of $\C$ and $\C$ is an instance of $\D$, no judgment
about the relationship of $\B$ to $\D$ can be inferred.

Another difference between subclassing and {\txt{instanceof}}
relationships is that cycles can occur among instance class
relationships. The simplest such cycle is:

\begin{verbatim}
class C kind C {}
\end{verbatim}

Although this pathology appears to be dangerously close to Russell's
paradox, it does not lead to inconsistencies.  Class \C\ is an
instance of itself. It can be used in any context where one of its
instances can be used.  \C\ can be thought of as a self-replicating
value. This class hierarchy causes no problems for method lookup
because method resolution never proceeds through more than a single
containing class to an instance class. If the programmer uses a field
of \C\ to initialize a class field of \C\, the evaulation of the
expression \C\ may fail to terminate.  But this is no different than
the possibility of non-termination from any other self-reference.
Therefore, we see no reason to disallow it.

@@

\begin{verbatim}
class Dimension { ... }
class Length kind Dimension { ... }
class Meter kind Length { ... }
\end{verbatim}

Here {\txt{Meter}} is a singleton class (there is only one Meter, in
the Platonic sense).  We have easily expressed the desired type
relationships, and can statically check program invariants that rely
on dimensional relationships.

@@

\begin{verbatim}
class Species { ... }
class Eagle kind Species {
      class Eagle make(String name) { ... }
}
Eagle harry = Eagle.make("Harry")
\end{verbatim}

@@

\begin{verbatim}
class Species { ... }
class TheEagle extends Species { ... }
class Eagle { ... }
Eagle harry = new Eagle("Harry")
\end{verbatim}

There are at least two substantial problems with this code.  First,
{\txt{TheEagle}} and {\txt{Eagle}} have no relationship to each other in
the type system.  One is a singleton class, representing a particular
species, and one is a name for a set of objects, being the members of
that species.  The fundamental nature of species has a two-level
containment relationship, which the type system fails to express.
This means that our type checker cannot determine what we want and
cannot help us avoid mistakes.

Another problem, which is in some ways just a symptom of the first
problem but which bears special attention, is the use of generic
types.  Consider the following class definition, added to the
definitions of \txt{Species}, \txt{theEagle}, etc. above.

\begin{verbatim}
class C<T extends X> {
   foo(T x) { return T.isEndangered() }
}
\end{verbatim}

If we want to perform the call {\txt{foo(harry)}}, then we have two
choices for the bound \txt{X} on {\txt T}.  It must be either \Object,
in which case {\txt{m}} must be a method that \Object\
understands. Otherwise, it must be \txt{Eagle}, in which case the
function cannot handle multiple {\txt{Species}}.  The ``solution'' in
conventional languages is to use a bound of \Object\ and insert a
downcast, which fails at runtime if the wrong argument is passed.
This is the problem that generics were intended to alleviate.  In
MCJ {\txt T} can be bound by kind {\txt{Species}}, and instantiated
with {\txt{Eagle}}, allowing the original call to be type-checked.
This gives the programmer both safety and expressiveness.

\paragraph{Metaclasses and Nested Classes}
Some of the problems are alleviated in a language with nested, or
inner classes.  In the language Scala\cite{Scala} the following
program expresses some of the desired invariants:

\begin{verbatim}
class Species {
  class Member {}
}
object Eagles extends Species;
class Eagle() extends Eagles.Member;
val harry = new Eagle;
\end{verbatim}
\footnote{This example, and this discussion, was suggested by an
anonymous reviewer.}

This program automates many of the invariants required in a language
without metaclasses.  However, two seperate classes must still be
constructed per \txt{Species}.  Further, the fundamental problem, of
not being able to group classes by functionality in the same way as
instances, still remains.  Inner classes are not constrained by
invariants imposed by their container.  In MCJ, the invariants of the
kind are enforced by the type system.

@@

One of the advantages of our metaclass framework is that it allows
clean expression of several common design patterns in the language,
rather than requiring that they be expressed with abstraction
techniques on top of the language.  For example, a number of the
classic design patterns deal with object creation, as discussed above
in section \ref{privconst}.  It is clear that the concept of a Factory
method is easily (and necessarily) expressed in MCJ.  However, a
number of other patterns also find easy expression.  The Abstract
Factory pattern is a simple application of inheritance in the
metaclass hierarchy.  In this, several classes could share a kind,
giving them all a single interface for construction. The Prototype
pattern is, as mentioned in the Design Patterns book, supported in
languages with metaclasses.  A class is in many ways a clone of its
metaclass.  Finally, there is no need for the Singleton pattern; a
class itself can serve as a singleton. If no factory methods are
provided, no instances can be created at runtime.  While classes can
be used as limited singletons in the Java Programming Language and in
\Cpp, doing so is undesirable because the static behavior of a class
cannot inherit from another class.  In fact, the inflexibility of
class or static operations is one of the rationales cited for the
Singleton pattern.

@@

A number of symbols are used to abbreviate keywords: $\superSym$
stands for \extends\ and $\kindSym$ stands for \kind. Also
$@$ represents concatenation of sequences of syntactic constructs.
Finally, $CT(\T)$ is a lookup in the class table for the definition of
$\T$.

A number of restrictions on MCJ programs are implicit in the formal
rules.  First, we assume that all sequences of methods and fields are
free of duplicates.  Second, there is an implicit well-formedness
constraint on programs that no class be a superclass of itself, either
directly or indirectly (however, as discussed below, cycles in the
kind hierarchy are allowed).  Third, we assume that \this is
never used as the name of a variable, method or field.  We also take
\Object\ to be a distinguished member of the hierarchy, with no
specific definition, which does not have a superclass or a kind.

@@

This is a class named $\C$, with no type arguments, with kind \Object\
and superclass \Object\, which has no fields or methods whatsoever.
In examples that follow, we omit empty type parameter lists.  Note
that both the kind and the superclass are $\Object$.  Unlike Smalltalk
or ObjVLisp, MCJ does not have a distinguished class
$\Class$.\footnote{This is not the same as the \Class\ class used in
reflection.}  Interestingly, we determined that a class $\Class$ would
need no consideration from the type system.  Because both $\Class$ and
$\Object$ would sit atop the hierarchy with no contents, there is no
need to include both of them.  In a more substantial language, where
$\Object$ might have methods or fields, there might be a use for a
distinguished $\Class$ class, but it would not need any special
treatment by the type system.

A more complicated MCJ class is:

\begin{verbatim}
class Pair<A extends Object kind Object,
           B extends Object kind Object>
{
 class Pair<A,B> make(A a, B b){ 
    return new <A,B> (a,b) 
} ;
 A fst;
 B snd;
 Pair<B,B> setfst(B b){
     return new <B,B> (b, this.snd);
 };
}
\end{verbatim}

This class specifies a polymorphic pair data structure.  It also
demonstrates a number of important MCJ features.  First, we see two
different kinds of methods.  The {\txt{make}} method creates a new
instance of {\txt{Pair}}.  This method contains a {\txt{new}}
expression, which initializes the instance fields of the class
positionally, so that the first argument to {\txt{make}} becomes
{\txt{fst}}, and the second becomes {\txt{snd}}.  Also, note that we
must mention the type parameters of the newly created instance.

Second, we have an instance method, {\txt{setfst}}, which creates a new
pair with the old second element, and a new first element which has
the same type as the second element.

Finally, adding the following to the above definition of {\txt{Pair}}
gives a full MCJ program:

\begin{verbatim}
class O
{
 class O make(){return new () ()};
}

@@

This program, which creates two instances of {\txt O}, then creates
a {\txt{Pair}} to hold them both, and finally selects one of the two to
become the value of the program, demonstrates three of the expression
forms in MCJ.  {\txt{Pair<O,O>}} is a type, which, by itself, can
be a value.  {\txt{O.make()}} is a method invocation (here with a
class, not an instance, as the receiver).  The entire expression is a
field access, of the {\txt{fst}} instance field.
